---
title: Examination Project in R
author: Marina Dimitriou
email: marina.dimitriou.8935@student.uu.se
date: 02 December 2024
mainfont: "Times New Roman"
sansfont: "Arial"
mathfont: "TeX Gyre Termes Math"
monofont: "Courier New"
fontsize: 12pt
output: 
 html_document: 
    code_folding: show
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 60))
```


# Setup environment
## Load libraries
```{r libraries}
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ncappc)
```


## Define functions
In this section, the functions that will be used later are defined.
```{r functions}
# Function to calculate AUC with the trapezoidal rule
## Calculate trapezoids for each interval [Trn=(tn-tn-1)*(Cn-1+Cn)/2]
## Add the the trapezoids together
## Add the residual area (Cn/k)
measure_auc <- function(time, conc, k) {
  trapezoids <- diff(time) * (head(conc, -1) + tail(conc, -1)) / 2
  auc <- sum(trapezoids)
  auc <- auc + conc[length(conc)] / k
  return(auc)
}

# Function to numerically summarize the PK variables 
num_summarize <- function(data, columns) {
  for (col in columns) {
    cat("\nNumerical Summary for", col, ":\n")
    stats <- data.frame(
      mean = mean(data[[col]]),
      median = median(data[[col]]),
      sd = sd(data[[col]]),
      range = diff(range(data[[col]])),
      min = min(data[[col]]),
      max = max(data[[col]])
      )
      print(stats)
  }
}
```

## Set working directory
```{r wd}
# Set the working directory for the entire notebook
knitr::opts_knit$set(root.dir = "C:/Users/mardi/Documents/Exam")
```

# Data management 
This section includes data import, variable assignment, dataset reorganisation (merge + long format)

```{r data management}
# Import the data sets

dir_name <- paste("Orig_data", sep = "/") # For path to the data

## Set the file names and file paths
file_name_pk <- "BPI889_PK_17.csv"
file_name_cov <- "BPI889_cov_17.txt"
file_path_pk <- paste(dir_name, file_name_pk, sep = "/")
file_path_cov <- paste(dir_name, file_name_cov, sep = "/")

## Import both datasets as R-objects
data_pk <- read.csv(file = file_path_pk, # It is a csv file
                    header = TRUE,    # The 1st row of csv has the column names
                    as.is = TRUE,     # Prevents turning character to factor
                    na.strings = ".") # The dots should be treated as NA

data_cov <- read.table(file = file_path_cov,  # It is a txt file
                       header = TRUE, # The 1st row of txt has the column names
                       sep = ",",     # The column separator is comma
                       as.is = TRUE)  # Prevents turning character to factor


## Visually inspect the imported datasets
str(data_pk)
str(data_cov)
head(data_pk, n = 10)  # To inspect the first 10 rows of data_pk
head(data_cov, n = 10) # To inspect the first 10 rows of data_cov

# Variable assignment and dataset reorganization

## Change the names of the columns of data_cov to make them more readable
names(data_cov) <- gsub("\\.$", "", names(data_cov))  # Remove last dot
names(data_cov) <- gsub("\\.+", "_", names(data_cov)) # Replace dots with "_"
names(data_cov) <- gsub("^X", "CYP", names(data_cov)) # Replace "X" with "CYP"
names(data_cov)[1] <- "pat" # Change the first name from patid to pat

names(data_cov) # To check the changed names

## Change data_pk from a wide to a long format
tidy_pk <- gather(data = data_pk, key = Time_h, value = Conc_mg_L, -pat)

nrow(tidy_pk) # To print the dimensions of tidy_pd
ncol(tidy_pk) # To print the dimensions of tidy_pd
head(tidy_pk, n = 5) # To output the first 5 rows of tidy_pd

## Format Numeric variables

### Concentration and time need to be numeric not characters
summary(tidy_pk$Conc_mg_L) # To check if variable is numeric
summary(tidy_pk$Time_h) # To check if variable is numeric

tidy_pk$Time_h <- gsub("[^0-9.]", "", tidy_pk$Time_h) # To remove letters
head(tidy_pk$Time_h) # Check if it was removed successfully 

unique(tidy_pk$Time_h) # To look at the unique values in Time_h
tidy_pk$Time_h <- as.numeric(tidy_pk$Time_h) # To convert Time_h to numeric
str(tidy_pk$Time_h) # To check if it was converted to numeric 

##Format categorical variables (Sex and CYP)

###Factorize Sex
data_cov$Sex <- factor(data_cov$Sex, levels = c("M", "F"),
                         labels = c("Male", "Female"))

### Define the column names with enzymes to factorize
cyp_columns <- c("CYP2D6", "CYP3A4", "CYP2C9", "CYP2C19")

### Factorize the enzymes with levels of metabolizing activity
for (col in cyp_columns) {
  data_cov[[col]] <- factor(data_cov[[col]], 
                            levels = c(0, 1, 2), 
                            labels = c("Poor", "Normal", "Extensive"))
}

str(data_cov) # To check if Sex and CYP columns were factorized successfully

## Merge the 2 data sets by patient ID ("pat")
data_all <- merge(tidy_pk, data_cov, by = "pat") 

## Reorder rows in the new data set by first increasing ID and then TIME
data_all <- data_all[order(data_all$pat, data_all$Time_h), ]
```

# Variable calculations
This section includes the **calculation of body size measurement (BFP)**, **categorization of BFP** and **PK variable calculation**

```{r variable calculations}
# Calculation of body size measurement

## Calculate BFP for men and women (height was in cm and divided by 100 to m)
data_all$BFP_percent <- ifelse(
  data_all$Sex == "Male", 
  1.2 * (data_all$Weight_kg / (data_all$Height_cm / 100)^2) +
    (0.23 * data_all$Age_yrs) - 16.2, 
  1.2 * (data_all$Weight_kg / (data_all$Height_cm / 100)^2) +
    (0.23 * data_all$Age_yrs) - 5.4
)

# Categorization of BFP

## Create a new variable CBFP to categorize BFP based on a 24% cutoff
### Assign 0 for BFP under 24% and 1 for BFP above 24%
data_all$CBFP <- ifelse(data_all$BFP < 24, 0, 1)

### Factorize CBFP                   
data_all$CBFP <- factor(data_all$CBFP,
                          levels = c(0, 1),
                          labels = c("Below 24%", "Above 24%"))

table(data_all$CBFP) # To summarize CBFP

str(data_all) # To see that CBFP was factorized correctly

# PK variable calculation

## Calculate Cmax (mg/L) and tmax (h) and add them in columns to the data_all
data_all <- data_all %>%
  group_by(pat) %>%
  mutate(
    Cmax_mg_L = max(Conc_mg_L, na.rm = TRUE),
    tmax_h = Time_h[which.max(Conc_mg_L)]
  ) %>%
  ungroup()

str(data_all) # To check the 2 new columns of containing Cmax and tmax

## Find terminal phase to calculate k and then t1/2

### Create spaghetti plot to display individual concentrations of BPI889 vs time
### To roughly check which time points should be included for the terminal phase
ggplot(data_all, aes(x = Time_h, y = Conc_mg_L, group = pat)) +
  geom_line() 

### Time=48h should be excluded after reviewing graph (not linear decline)
### Check if all time points after tmax and without 48h, are the terminal phase 
### Perform linear regression on the log-transformed concentration for each pat
### Threshold for log-linearity for R-squared is 0.9

check_terminal <- data_all %>%
  filter(!is.na(Conc_mg_L)) %>%
  filter(Time_h > tmax_h & Time_h != 48) %>%
  group_by(pat) %>%
  summarise(
    r_squared = ifelse(n() < 2, NA, summary(lm(log(Conc_mg_L) ~ Time_h))$r.squared),
    is_terminal = r_squared > 0.9 
  )

### Find if there are patients without enough time points for linear regression
any_na <- any(is.na(check_terminal$is_terminal))
### Count the number of patients who do not meet the log-linearity threshold
false_count <- sum(check_terminal$is_terminal == FALSE)

### Print a message if all patients meet the log-linearity requirement or not
terminal_approval <- ifelse(
  false_count == 0 & any_na == FALSE, 
  "This is an acceptable terminal phase for all patients",
  "This isn't an acceptable terminal phase for all patients.")
print(terminal_approval) # Print the message

## Calculate k and then calculate half life (t_half_h)
### Fit the regression model and store in a list
### Calculate k from the slope (negate to make it positive)
k_values <- data_all %>%
  group_by(pat) %>%                                 
  filter(Time_h > tmax_h & Time_h != 48) %>%         
  summarise(
    model = list(lm(log(Conc_mg_L) ~ Time_h)),  
    k = -coef(model[[1]])[2],  
    t_half_h = log(2) / k                  
  ) %>%
  ungroup()

### Add k and half life (t_half_h) as a column to data_all
data_all <- merge(data_all, k_values[, c("pat", "k", "t_half_h")], by = "pat")


## Calculate AUC for each patient using the function 
auc_values <- data_all %>%
  group_by(pat) %>%                         
  filter(!is.na(Conc_mg_L)) %>%             
  summarise(AUC_mgh_L = measure_auc(Time_h, Conc_mg_L, k = k[1])) %>%
  ungroup() 

### Add AUC (AUC_mgh_L) to data_all
data_all <- merge(data_all, auc_values[, c("pat", "AUC_mgh_L")], by = "pat")

## Calculate CL for each patient and add it as a column (CL_L_h) to data_all
data_all$Dose_mg <- 200 # Add column to data_all for the dose which was 200mg
data_all$CL_L_h <- data_all$Dose_mg/data_all$AUC_mgh_L  # Calculate CL=Dose/AUC

str(data_all) # To check the new columns
```


# Data exploration
This section includes the **numerical summary of PK variables** and the **graphical assessment of** 1) PK profiles, 2) PK variable correlations, 3) PK variable-CYP enzyme correlations, 4) PK variable-body size measurement correlation with linear regression


```{r data exploration}
# Numerical summary of PK variables

## Create data_first that has only the 1st obs for each patient
### Because the PK variables are the same for every row of each patient
data_first <- data_all[!duplicated(data_all$pat), ]

## Create vector that has the names of the columns of PK variables
PK_variables <- c("Cmax_mg_L", "tmax_h", "t_half_h", "CL_L_h")

## Use the function to numerically summarize the PK variables
num_summarize(data_first, PK_variables)

# Graphical assessment

## Create spaghetti plot to display individual concentrations of BPI889 vs time

data_clean <- data_all %>% filter(!is.na(Conc_mg_L)) # need to remove NA to plot

ggplot(data_clean, 
       aes(x = Time_h, y = Conc_mg_L, group = pat, color = as.factor(pat))
  ) +
  geom_line() +
  scale_x_continuous(
    breaks = c(0.5, 1, 2, 4, 6, 8, 12, 24 , 48),
    labels = c("0.5", "1", "2", "4", "6", "8", "12", "24", "48")
  ) +
  labs(
    title = "Individual concentrations of BPI889 vs Time",
    x = "Time (hours)",
    y = "Concentration (mg/L)"
  ) +
  theme_minimal() + 
  theme(legend.position = "none") # Hide the legend because there are many pat

## Create a correlation plot between Cmax, tmax, t1/2, and CL
GGally::ggpairs(data_first[, PK_variables])

## Create box-and-whisker plots to display t1/2 and CL vs all the given enzymes
### Create first empty lists to store the plots
thalf_plots <- list()
cl_plots <- list()

### Loop through enzymes and create the box plots for t1/2
for (enzyme in cyp_columns) {
  tplot <- ggplot(data = data_first, aes(x = .data[[enzyme]], 
                  y = t_half_h, fill = .data[[enzyme]])) +
          geom_boxplot(outlier.colour = "black", outlier.shape = 16, 
                       outlier.size = 2, notch = FALSE) +
          labs(title = paste("Box plot of t1/2 vs", enzyme, "Enzyme"),
                y = "t1/2 (h)") +
          scale_fill_brewer(palette = "Set2") +
          theme_classic()
  
  thalf_plots[[enzyme]] <- tplot
}

### Display the t1/2 vs enzyme box plots in the desired way
ggpubr::ggarrange(plotlist = thalf_plots, ncol = 2, nrow = 2) 

### Loop through enzymes and create the box plots for clearance 
for (enzyme in cyp_columns) {
  clplot <- ggplot(data = data_first, aes(x = .data[[enzyme]], 
                    y = CL_L_h, fill = .data[[enzyme]])) +
            geom_boxplot(outlier.colour = "black", outlier.shape = 16, 
                         outlier.size = 2, notch = FALSE) +
            labs(title = paste("Box plot of CL vs", enzyme, "Enzyme"),
                 y = "Clearance (L/h)") +
            scale_fill_brewer(palette = "Set2") +
            theme_classic()
  
  cl_plots[[enzyme]] <- clplot
}

### Display the clearance vs enzyme box plots in the desired way
ggpubr::ggarrange(plotlist = cl_plots, ncol = 2, nrow = 2) 

## Create a scatter plot to display correlation between t1/2 and BFP, with lm

ggplot(data_first, aes(x = BFP_percent, y = t_half_h, color = CBFP )) +
  geom_point() +
  geom_smooth(method = "lm", color = "red" ) +
  labs(
    title = "Scatter Plot of t1/2 vs BFP",
    x = "Body Fat Percentage (%)",
    y = "t1/2 (h)"
  )
```

# Statistical testing
This section includes **ANOVA of the PK variables for the four individual CYP-enzymes** and the **linear regression of t1/2 versus BFP**.

```{r statistical testing}
# Perform an ANOVA on Cmax and CL across the four individual CYP-enzymes 
## The phenotypes of the CYPs are categorical
### as.formula is used in order for the output to have CYP name as row name
#### Οtherwise, it could be done without using as.formula

for (enzyme in cyp_columns) {
  # Perform ANOVA for Cmax
  formula_cmax <- as.formula(paste("Cmax_mg_L ~", enzyme))
  print(paste("ANOVA for Cmax and", enzyme))
  model_cmax <- aov(formula_cmax, data = data_first)
  print(summary(model_cmax))
  
  # Perform ANOVA for CL
  formula_cl <- as.formula(paste("CL_L_h ~", enzyme))
  print(paste("ANOVA for CL and", enzyme))
  model_cl <- aov(formula_cl, data = data_first)
  print(summary(model_cl))
}

# Perform a linear regression of t1/2 versus BFP.
lm_thalf_BFP <- lm(t_half_h ~ BFP_percent, data = data_first)

## Show the summary of the regression model
summary(lm_thalf_BFP)

### Extract p-value
p_value <- summary(lm_thalf_BFP)$coefficients[2, 4]

### Report if the p-value is significant
report_lm <- ifelse(
  p_value < 0.05,
  "The relationship between BFP and t1/2 is statistically significant.",
  "The relationship between BFP and t1/2 isn't statistically significant."
)
print(report_lm)
```
# Final answers to the research questions
From the statistical analysis performed in the section above, the distribution of BPI889 seems to be dependent on BFP and the elimination of BPI889 seems to be dependent on CYP2D6.